Geographically aware estimates of remotely sensed water properties for Chesapeake Bay

### 2022-09-13 manuscript/manuscript.tex
format and include rmse table
text edits

Chesapeake Bay is located in the Mid-Atlantic region of the United States and serves as the outlet of 9 major tributary systems. Of these, the Susquehanna River provides the majority of the freshwater supply to the Bay (\texttildelow$48\%$, Xu et al. 2012). The Bay is fairly shallow ($<$ 30m, \citet{murphyComparisonSpatialInterpolation2010}) and the hydrodynamics of the Bay are such that it is considered stratified and partially mixed (Xu et al. 2012).

@@

Prior data-driven models (e.g. \citet{urquhartGeospatialInterpolationMODISderived2013, vogelAssessingSatelliteSea2016}) of Bay salinity used Generalized Linear Models (GLM), Neural Networks (NN), Generalized Additive Models (GAMs), and Random Forests (RF). We chose to use the Random Forest algorithm in part because \citet{urquhartGeospatialInterpolationMODISderived2013} found that it produced nearly equivalent results to more complex methods (particularly NN) but also because of its ease of implementation and interpretation. We fit RF models using the Python package scikit-learn \citep{pedregosaScikitlearnMachineLearning2011} where each model was initialized with 250 trees, had a maximum depth of 20 tree “layers”, and was subjected to recursive feature elimination to produce the most parsimonious model. We subsequently “tuned” each model using k-fold cross validation to optimize the size and splitting criteria of each tree.

@@

We compiled over two decades of observational data (2000-2021) on salinity, temperature, and turbidity. Although this data came from a wide variety of sources, almost $80\%$ of the total unique observations came from three programs NDBC/CBIBS (\url{https://buoybay.noaa.gov/}), Maryland Continuous Monitoring Program (MDNR), and the Virginia Estuarine and Coastal Observing System (\url{http://vecos.vims.edu}). Because Chesapeake Bay has a mean optical depth of 0.89m \citep{urquhartRemotelySensedEstimates2012}, we filtered collected observations to those occurring in the top 1m of the water column. We further limited observations to those that had a timestamp during the MODIS overpass window time-of-day (approximately 1700 - 1900 hours). Finally, we implemented spatial and temporal averaging of observational data to a daily mean within each MODIS pixel. We supplemented observational data with a grid of “freshwater influence” (FWI) estimates where each grid cell represents the average of flow weighted distance of that cell from the 5 major tributaries (Susquehanna, James, Pautexent, Choptank, and Potomac). Rather than straight-line distances we used “in-water” distances calculated using the scikit-image Python package \citep{vanderwaltScikitimageImageProcessing2014}. We combined the FWI estimates with the pixel-wise daily mean observations and regressed them in RF models against data from the 500m resolution MODIS Aqua product, which provides 8 bands values for each pixel spanning from 405 to 877nm wavelength \citep{vermoteericMOD09GAMODISTerra2015}.

@@

We validated RF model predictions both quantitatively against a hold-out test set of observational data (one third of the total data) and qualitatively against data from the Chesapeake Bay Operational Forecast System (CBOFS, \citet{lanerolle2011second}). CBOFS provides data on Bay-wide water level, temperature, and salinity every 6 hours. The model itself uses National Oceanic and Atmospheric Admistration (NOAA) water level, salinity observations, and US Geological Survey (USGS) river discharge data to calibrate an implementation of the Rutgers University Regional Ocean Modeling System (ROMS, \citet{shchepetkinRegionalOceanicModeling2005}).

@@

>>>>

Chesapeake Bay is located in the Mid-Atlantic region of the United States and serves as the outlet of 9 major tributary systems. Of these, the Susquehanna River provides the majority of the freshwater supply to the Bay (\texttildelow$48\%$, \citet{xuClimateForcingSalinity2012}). The Bay is fairly shallow ($<$ 30m, \citet{murphyComparisonSpatialInterpolation2010}) and the hydrodynamics of the Bay are such that it is considered stratified and partially mixed \citep{xuClimateForcingSalinity2012}.

@@

Prior data-driven models (e.g. \citet{urquhartGeospatialInterpolationMODISderived2013, vogelAssessingSatelliteSea2016}) of Bay salinity used Generalized Linear Models (GLM), Neural Networks (NN), Generalized Additive Models (GAMs), and Random Forests (RF). We chose to use the Random Forest algorithm in part because \citet{urquhartGeospatialInterpolationMODISderived2013} found that it produced nearly equivalent results to more complex methods (particularly NN) but also because of its ease of implementation and interpretation. We fit RF models using the scikit-learn Python package \citep{pedregosaScikitlearnMachineLearning2011} where each model was initialized with 250 trees, had a maximum depth of 20 tree “layers”, and was subjected to recursive feature elimination to produce the most parsimonious model. We subsequently “tuned” each model using k-fold cross validation to optimize the size and splitting criteria of each tree. We evaluated the “importance” of a particular variable as the average impurity decrease of tree splits based on that variable \citep{pedregosaScikitlearnMachineLearning2011}.

@@

We compiled over two decades of observational data (2000-2021) on salinity (n=6k), temperature (n=10k), and turbidity (n=5k, Table S1). Although this data came from a wide variety of sources (n=20, Table S1), almost $80\%$ of the total unique observations came from three programs NDBC/CBIBS (\url{https://buoybay.noaa.gov/}), Maryland Continuous Monitoring Program (MDNR), and the Virginia Estuarine and Coastal Observing System (\url{http://vecos.vims.edu}). Because Chesapeake Bay has a mean optical depth of 0.89m \citep{urquhartRemotelySensedEstimates2012}, we filtered collected observations to those occurring in the top 1m of the water column. We further limited observations to those that had a timestamp during the MODIS overpass window time-of-day (approximately 1700 - 1900 hours). Finally, we implemented spatial and temporal averaging of observational data to a daily mean within each MODIS pixel. We supplemented observational data with a grid of “freshwater influence” (FWI) estimates where each grid cell represents the average of flow weighted distance of that cell from the 5 major tributaries (Susquehanna, James, Pautexent, Choptank, and Potomac). Rather than straight-line distances we used “in-water” distances calculated using the scikit-image Python package \citep{vanderwaltScikitimageImageProcessing2014}. We combined the FWI estimates with the pixel-wise daily mean observations and regressed them in RF models against data from the 500m resolution MODIS Aqua product, which provides 8 bands values for each pixel spanning from 405 to 877nm wavelength \citep{vermoteericMOD09GAMODISTerra2015}.

@@

We validated RF model predictions both quantitatively against a hold-out test set of observational data (one third of the total data) and qualitatively against data from the Chesapeake Bay Operational Forecast System (CBOFS, \citet{lanerolle2011second}). CBOFS provides data on Bay-wide water level, temperature, and salinity every 6 hours. The model itself uses National Oceanic and Atmospheric Administration (NOAA) water level, salinity observations, and US Geological Survey (USGS) river discharge data to calibrate an implementation of the Rutgers University Regional Ocean Modeling System (ROMS, \citet{shchepetkinRegionalOceanicModeling2005}).

@@

\begin{figure}[ht!]
    \begin{center}
          \includegraphics[width=0.75\textwidth,keepaspectratio]{figures/rf_stats_table_1_}          
    \end{center}    
\end{figure}
